Interpretable Casual Interventions \\ Discourage Spurious Learning

### 2025-03-24 paper/main.tex
added paper section for stricter core risk minimization objective formulation, plus minor typos and rephrasing

We demonstrate that the normalizing behavior of softmax during the forward pass introduces conditions where an interpretation obtained from HiResCAMs fails to convey the true factors that contribute towards a prediction. To correct this, we propose an interpretability technique -- \textbf{ContrastiveCAMs} -- which reconstructs provably faithful activation mappings in the multiclass setting.

@@

Using ContrastiveCAMs, we also argue that Cross-Entropy Loss \citep{cox1958regression, mao2023cross} \textbf{motivates learning spurious correlations}, especially in settings where the target represents a small portion of the image (Section \ref{ce-motiv-spurious}). We further introduce an interpretable training objective with ContrastiveCAMs, which is equivalent to Core Risk Minimization \citep{singla2022core}.

@@

\subsection{Defining the Contrastive Optimization Objective}

@@

We therefore derive an optimization objective that attempts to maximize the contrast between the two activations maps. For the binary classification case, $(c, c')$ reflects that correct and incorrect class respectively.
\begin{gather}
	\max_{\theta} s_c - s_{c'} \equiv \max_{\theta} \sum^{D_1,D_2}_{i,j} \tilde{\mathcal{\bm{A}}}_{c,i,j}^{\text{HiResCAM}} + b_c - \sum^{D_1,D_2}_{i,j} \tilde{\mathcal{\bm{A}}}_{c',i,j}^{\text{HiResCAM}} - b_{c'}
\end{gather}
Ignoring the bias terms, our new optimization objective is:
\begin{gather}
	\max_{\theta} \sum^{D_1,D_2}_{i,j} \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, c'), i, j} ~~\text{ where }~~ \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, c')} := \tilde{\mathcal{\bm{A}}}_{c}^{\text{HiResCAM}} - \tilde{\mathcal{\bm{A}}}_{c'}^{\text{HiResCAM}}
\end{gather}
This is equivalent to the original ERM optimization objective, with the difference being the resultant values preserving spatial information in the form of the activation map instead of flattening the feature map. If $\tilde{\mathcal{\bm{A}}}_{c}^{\text{HiResCAM}} - \tilde{\mathcal{\bm{A}}}_{c'}^{\text{HiResCAM}} > 0$, the classification is correct -- if not, the classification is incorrect.

@@

To ensure that the model accurately predicts the target class, each contrastive comparison should yield a positive result. Therefore, the multiclass optimization objective under this paradigm is as follows:
\begin{gather}
	\max_{\theta} \sum^{D_1,D_2}_{j,k}\tilde{\mathcal{\bm{A}}}_{(c, c'),j,k}^{\text{contrastive}}\ \forall c' \in \mathbb{Z}_{+}(|C| - 1)
\end{gather}

@@

The primary distinction between the contrastive optimization objective over standard risk is that it preserves spatial information. We know at a pointwise level how each feature contributes to the final prediction.

We can use this to constrain the way optimization occurs -- namely, we only want the model to increase the likelihood of an object belonging to a certain class by actually looking at the features of the class, and not features of spurious correlation like the environment in which the object is present. The model through the process of training should learn to distinguish and subsequently ignore the background in determining the final prediction.

% TODO: formal analysis -- empirical performance test on scale augmentation + norm-change in feature representation.

In addition to learning invariances to spurious correlation, having information about the specific region containing the classification target also motivates learning scale-robust representations of the target.

@@

\subsection{Training Detail}

@@

\begin{table}[H]
	\centering
	\caption{Accuracies under various ablations, Relative Foreground Sensitivity, on FineTuned ResNet-50 models trained using varying approaches.}
	\begin{tabular}{c|c|ccc|c}
		\toprule
		\textbf{Method} & \textbf{None} $(\uparrow)$ & \textbf{Gray Mask FG} $(\downarrow)$ & \textbf{Gray BBOX FG} $(\downarrow)$ & \textbf{Tile FG} $(\downarrow)$ & \textbf{RFS} $(\uparrow)$ \\
		\midrule
		Cross-Entropy & 94.25 & 75.94 & 69.39 & 67.38 & -0.18 \\
		\midrule
		CORM & 92.91 & 76.20 & 69.12 & 68.32 & -0.08 \\
		DRF & \bf 94.39 & 73.53 & 67.51 & 66.71 & -0.27 \\
		CORM + DFR & 91.31 & 72.59 & 63.64 & 63.90 & -0.23 \\
		\midrule
		ACE (Ours) & 89.20 & 47.33 & 34.27 & 34.93 & 0.21 \\
		ACE + SWD (Ours) & 91.20 & \bf 36.13 & \bf 27.20 & \bf 31.60 & \bf 0.28 \\
		ACE + SWD + RevUPs (Ours) & 75.86 & \bf 10.40 & \bf 09.86 & \bf 13.19 & \bf 0.57 \\
		\bottomrule
	\end{tabular}
\end{table}

>>>>

We demonstrate that the normalizing behavior of softmax during the forward pass introduces conditions where an interpretation obtained from HiResCAMs fails to convey the true factors that contribute towards a prediction. To correct this, we propose an interpretability technique -- \textbf{ContrastiveCAMs} (Section \ref{sec:define-cont-cams}) -- which reconstructs provably faithful activation mappings in the multiclass setting.

@@

Using ContrastiveCAMs, we also argue that Cross-Entropy Loss \citep{cox1958regression, mao2023cross} \textbf{motivates learning spurious correlations}, especially in settings where the target represents a small portion of the image (Section \ref{ce-motiv-spurious}). We further introduce an interpretable training objective with ContrastiveCAMs, which is a stricter counterpart of Core Risk Minimization \citep{singla2022core} (Section \ref{sec:constraining-optimization}).

@@

\subsection{Defining the Contrastive Optimization Objective}
\label{sec:define-cont-cams}

@@

We therefore derive an optimization objective that attempts to maximize the contrast between the two activations maps. For the binary classification case, $(c, c')$ reflects that correct and incorrect class respectively.
\begin{gather}
	\max_{\theta} s_c - s_{c'} \equiv \max_{\theta} \sum^{D_1,D_2}_{i,j} \tilde{\mathcal{\bm{A}}}_{c,i,j}^{\text{HiResCAM}} + b_c - \sum^{D_1,D_2}_{i,j} \tilde{\mathcal{\bm{A}}}_{c',i,j}^{\text{HiResCAM}} - b_{c'}
\end{gather}
Ignoring the bias terms, the new optimization objective becomes:
\begin{gather}
	\max_{\theta} \sum^{D_1,D_2}_{i,j} \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, c'), i, j} ~~\text{ where }~~ \tilde{\mathcal{\bm{A}}}^{\text{contrastive}}_{(c, c')} := \tilde{\mathcal{\bm{A}}}_{c}^{\text{HiResCAM}} - \tilde{\mathcal{\bm{A}}}_{c'}^{\text{HiResCAM}}
\end{gather}
This is equivalent to the original ERM optimization objective, with the difference being the resultant values preserving spatial information in the form of the activation map instead of flattening the feature map. If $\tilde{\mathcal{\bm{A}}}_{c}^{\text{HiResCAM}} - \tilde{\mathcal{\bm{A}}}_{c'}^{\text{HiResCAM}} > 0$, the classification is correct -- if not, the classification is incorrect.

@@

To ensure that the model accurately predicts the target class, each contrastive comparison should yield a positive result. Therefore, the multiclass optimization objective under this construction reads:
\begin{gather}
	\max_{\theta} \sum^{D_1,D_2}_{j,k}\tilde{\mathcal{\bm{A}}}_{(c, c'),j,k}^{\text{contrastive}}\ \forall c' \in \mathbb{Z}_{+}(|C| - 1)
\end{gather}

@@

The primary distinction between the contrastive construction over standard risk is that it preserves spatial information. We know at a pointwise level how each feature within the contrastive map contributes to the final prediction.

\paragraph{Core Risk Minimization \citep{singla2022core}} proposes a generalization of Emprical Risk Minimization, the objective is to minimize prediction error while non-core regions are injected with gaussian noise:
\begin{equation}
	\min_\theta \mathbb{E}_{(\bm{x}, \bm{h}, y)} [ \mathbb{E}_{\bm{z} \sim \mathcal{N}(0, \sigma^2 \bm{I})} \ell(f_\theta(\bm{x} + \bm{z} \odot (1 - \bm{h})), y) ]
\end{equation}
Here $\ell$ represents cross-entropy, $\bm{x}$ represents the input, $\bm{h}$ is the set of core features corresponding to the input encoded as a binary mask, $y$ is the label corresponding to the input and $f_\theta$ is the parameterized network.

We can extend this to constrain the way optimization occurs. In principle, the objective is consistent -- we only want the model to increase the likelihood of an object belonging to a certain class by actually looking at the features of the class, and not features of spurious correlation like the environment in which the object is present.

The model should, through the process of training learn to distinguish and subsequently ignore the background in determining the final prediction. Analogous to \citep{singla2022core}, we define the core-constrained optimization objective as follows:
\begin{equation}\label{icrm}
	\min_\theta \mathbb{E}_{(\bm{x}, \bm{h}, y)} \left[ \int_{\mathbb{R}^{C \times D_1 \times D_2}} \ell(f_\theta(\bm{x} + \bm{z} \odot (1 - \bm{h})), y)~d\bm{z} \right]
\end{equation}

For the experiments in this paper, we consider the segmentation mask of the image as a proxy for core features that causally determine the prediction. However, the same framework can be used to apply constrain learning over an arbitrary mask $\bm{h}$ (and subsequently invariance over mask $1 - \bm{h}$), which can vary depending on the task.

% TODO: formal analysis -- empirical performance test on scale augmentation + norm-change in feature representation.
%In addition to learning invariances to spurious correlation, having information about the specific region containing the classification target motivates learning scale-robust representations of the target.

In the following subsections, we outline cost functions designed to minimize (\ref{icrm}) using contrastive maps to directly learn invariance to spurious features while generating contrast over the target region.

@@

\subsection{Training Details}

@@

\begin{table}[H]
	\centering
	\caption{Accuracies under various ablations, Relative Foreground Sensitivity, on FineTuned ResNet-50 models trained using varying approaches.}
	\begin{tabular}{c|c|ccc|c}
		\toprule
		\textbf{Method} & \textbf{None} $(\uparrow)$ & \textbf{Gray Mask FG} $(\downarrow)$ & \textbf{Gray BBOX FG} $(\downarrow)$ & \textbf{Tile FG} $(\downarrow)$ & \textbf{RFS} $(\uparrow)$ \\
		\midrule
		Cross-Entropy & 94.25 & 75.94 & 69.39 & 67.38 & -0.18 \\
		\midrule
		CORM & 92.91 & 76.20 & 69.12 & 68.32 & -0.08 \\
		DFR & \bf 94.39 & 73.53 & 67.51 & 66.71 & -0.27 \\
		CORM + DFR & 91.31 & 72.59 & 63.64 & 63.90 & -0.23 \\
		\midrule
		ACE (Ours) & 89.20 & 47.33 & 34.27 & 34.93 & 0.21 \\
		ACE + SWD (Ours) & 91.20 & \bf 36.13 & \bf 27.20 & \bf 31.60 & \bf 0.28 \\
		ACE + SWD + RevUPs (Ours) & 75.86 & \bf 10.40 & \bf 09.86 & \bf 13.19 & \bf 0.57 \\
		\bottomrule
	\end{tabular}
\end{table}
